---
title: "R and Databases"
author: '@lachlandeer'
date: "August 22, 2017"
output: 
    html_document:
        theme: yeti
        highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the Required Packages

Let's load the packages we need to connect a database to R, and use `dplyr` to interact with the database

```{r load_packages}
library(dbplyr)
library(dplyr)
```

## Connecting to the Auctions Data

The SQLite database is contained in a single file `../database/auctions_data.db`. 
We connect to it using the function `src_sqlite`

```{r load_database}
auctions_db <- src_sqlite("../database/auctions_data.db")
```

The `src_sqlite()` command does not load the data into the R session (as the `read.csv()` function did). Instead, it merely instructs R to connect to the SQLite database contained in the `auctions_data.db` file.

(You can use the `src_mysql()`, `src_postgres()` and `src_bigquery()` to connect to the other database types supported by dplyr.)

Let’s take a closer look at the database we just connected to:

```{r}
auctions_db
```

So R correctly sees that we have the three tables we have previously been working with.

## Querying with SQL syntax

To connect to tables within a database, you can use the tbl() function from dplyr. This function can be used to send SQL queries to the database. To demonstrate this functionality, let’s select the columns "AuctionsID", "Volume", and "District" from the surveys table:

```{r}
tbl(auctions_db, sql("SELECT AuctionID, Volume, District FROM auctions"))
```

## Querying the database with the dplyr syntax

One of the strengths of dplyr is that the same operation can be done using dplyr's verbs instead of writing SQL. First, we select the table on which to do the operations by creating the auctions object, and then we use the standard dplyr syntax as if it were a data frame:

```{r}
auctions <- tbl(auctions_db, "auctions")

auctions %>%
    select(AuctionID, Volume, District)
```

In this case, the auctions object behaves like a data frame. Several functions that can be used with data frames can also be used on tables from a database. For instance, the head() function can be used to check the first 10 rows of a table:

```{r}
bids <- tbl(auctions_db, "bids")

head(bids, n=10)
```

Some functions don’t work quite as expected. For instance, let’s check how many rows there are in total using `nrow()`:

```{r}
nrow(bids)
```

You might have already noticed that the first line of the head() output included ?? indicating that the number of rows wasn’t known.

The reason for this behavior highlights a key difference between using `dplyr` on datasets in memory (e.g. loaded into your R session via read.csv()) and those provided by a database. To understand it, we take a closer look at how dplyr communicates with our SQLite database.

## SQL Translation

Behind the scenes, `dplyr`:

1. translates your R code into SQL
2. submits it to the database
3. translates the database’s response into an R data frame

To lift the curtain, we can use dplyr's `show_query()` function to show which SQL commands are actually sent to the database:

```{r}
show_query(head(bids, n=10))
```
Instead of having to formulate the SQL query ourselves - and having to mentally switch back and forth between R and SQL syntax - we can delegate this translation to dplyr.

dplyr, in turn, doesn’t do the real work of subsetting the table, either. Instead, it merely sends the query to the database, waits for its response and returns it to us.

That way, R never gets to see the full surveys table - and that’s why it could not tell us how many rows it contains. On the bright side, this allows us to work with large datasets - even too large to fit into our computer’s memory.

dplyr can translate many different query types into SQL allowing us to, e.g., select() specific columns, filter() rows, or join tables.

To see this in action, let’s compose a few queries with dplyr.

## Simple Database queries

Let's reproduce one of the queries we wrote directly using SQL syntax earlier in the module. We will select all bids from bidders 1 and 4, and keep only the bid, bidderID and auctionID:

```{r}
bids %>%
    filter(bidderID %in%  c(1,4)) %>%
    select(Bid, BidderID, AuctionID) %>%
    arrange(BidderID, AuctionID)
```

Or we can filter bids for bidders that have a BidderID that's and even number:

```{r}
bids %>%
    filter(bidderID %% 2 == 0) %>%
    select(Bid, BidderID, AuctionID) %>%
    arrange(BidderID, AuctionID)
```

 why are there only 10 rows?

The last line:

```{r, eval=FALSE}
## # ... with more rows
```

indicates that there are more results that fit our filtering criterion. Why was R lazy and only retrieved 10 of them?
Laziness

Hadley Wickham, the author of dplyr explains:

> When working with databases, **`dplyr`** tries to be as lazy as possible:
>
> * It never pulls data into R unless you explicitly ask for it.
> * It delays doing any work until the last possible moment - it collects together
> everything you want to do and then sends it to the database in one step.

To instruct R to stop being lazy, e.g. to retrieve all of the query results from the database, we add the `collect()` command to our pipe. 
It indicates that our database query is finished: time to get the _final_ results and load them into the R session.


```{r}
subset_bids <- bids %>%
                filter(bidderID %% 2 == 0) %>%
                select(Bid, BidderID, AuctionID) %>%
                arrange(BidderID, AuctionID) %>%
                collect()
```

Now we have all 17 rows that match our query in a `data.frame` and can continue to work with them exclusively in R, without communicating with the database.

```{r}
str(subset_bids)
```

## SQL Join with `dplyr`

We can use the dplyr `inner_join` function to combine data from two tables.
Let's again collect the minimum, maximum and average bid for each bidder, and link this to the bidder's name:

```{r}
bidders <- tbl(auctions_db, "bidders")

bids %>%
    group_by(BidderID) %>%
    summarise(smallestBid = min(bid,  na.rm = TRUE),
              averageBid  = mean(bid, na.rm = TRUE),
              largestBid  = max(bid,  na.rm = TRUE) 
              ) %>%
    inner_join(bidders) %>%
    select(FirstName, LastName, smallestBid, averageBid, largestBid) %>%
    collect()
```